Extended Early Adulthood and Attitude Change

### 2024-03-29 manuscript/formative_period.Rmd
kkedits329

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed (Elder 2018). Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These models of social imprinting have implications for the kinds of patterns we should observe in peoples' attitudes over time, and suggest specific forms that opinion change should take. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality traits [@wagner2019]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort -- the year of a person's birth - explains much more variance than period -- the year of observation -- for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

## Biological Mechanisms:

The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and science of how people process information. This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate (Bartels and Jackman 2014; Sherratt and Morand-Ferron 2018; Gopnik 2020). However, because people have already encountered a lot of information by the time they reach adulthood, each new piece has less of an impact on their overall dispositions [@bartels2014]. The distinctive thing about our early years, then, is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life (Mata and von Helversen 2015). Supporting this model is evidence that early formative periods have been observed not just in people, but across species. In a study of social learning among birds, for instance, Aplin et al (2017) show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

@@

## Life Course and Social Roles:

@@

Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how we have chosen to structure the life course, or the interaction between these life course stages and biological mechanisms. Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological factors in explaining why we observe "impressionable years."

@@

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood. In the United States, both the median age at first marriage and median age of first child birth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring later in the life-course and potentially in different orders.

If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect in how long exploration, and therefore durable change, lasts throughout the life-course. If this is the case, we should expect to see lower levels of attitude stability (and therefore more change) among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born say in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect that people aged 25-33, who in earlier cohorts displayed opinion patterns similar to middle-aged respondents [@krosnic1991; @alwin1991], should more closely resemble the youngest adults in more recent years. This produces our first hypothesis:

@@

## Changing Impressionable Years:

A second trend in early adulthood relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld [-@rosenfeld2007] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages in the United States is that young adults are increasingly escaping the controlling gaze of their parents. When young people live with, or are financially dependent on, their parents, the latter have more say about their children's choice of partner. Social and cultural changes that have increased independence from parental authority and mobility mean they have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.

Rosenfeld's work suggests that the social conditions where we spend our impressionable years might shape how much exploration we can engage in. An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration. This assertion in turn engenders a prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.

@@

Reliability reflects not just the degree to which people make errors , but the baseline level of variance around a fixed attitude position that people express.

>>>>

The idea that the early stages of the life course are especially influential in the process of attitude formation is dominant across contemporary accounts of socialization [@dinas2014; @guhin2021]. The principal argument across a range of perspectives is that the circumstances in which people are brought up leave an enduring and traceable mark on their personal culture -- worldviews, attitudes, dispositions, behaviors, and preferences [@gerber1998]. In the study of political socialization, this perspective enjoys a longstanding history. "Generational effects" are used to explain the development of political preferences, longitudinal shifts in attitudes, and differences across age-groups [@mannheim1952; @bartels2014]. However, the notion of "impressionable years" is more general than just political preferences. The concept of cohort effects -- widespread in sociology -- rests on the assumption that peoples' early years play key role in shaping not just political opinions, but peoples' cognition on many dimensions, shaping their choices and life outcomes well beyond the time when this period has passed [@elder2016]. Similarly, practice theories associated with Bourdieu suggest that the "past conditions of production" durably shape peoples' cognition, reactions, and behavior well into adulthood [@bourdieu1990]. Thus, the notion of a formative period at the early stages of the life-course underpins diverse and longstanding theories of behavior.

These models of social imprinting have implications for the kinds of patterns we should observe in a population's attitudes over time and suggest specific forms that within-person opinion change should take over time. The "impressionable years" hypothesis -- that people demonstrate more opinion change early in life followed by a period of relative stability, especially in middle age -- has often been tested against other theories of opinion change over the life course. The most common alternatives are an "aging stability" or "increasing persistence" hypothesis that argues that openness to change gradually declines over time and a "life stages" hypothesis, in which people are open to attitude change both early in life and later in life, with stability principally happening in middle ages [@visser2004a]. While different attitudes occasionally follow different patterns, the impressionable years hypothesis has generally been shown to be true for broad political ideology, partisan identification, and specific policy preferences [@krosnick1989; @alwin1991; @sears1999]; social attitudes [@danigelis2007]; and interest in politics [@prior2010]; as well as things like personality traits [@wagner2019]. More generally, recent research has found that, with a few notable exceptions of high-salience events such as national shifts around gay rights, durable change of social opinion is quite rare among U.S. adults [@kiley2020], and Vaisey and Lizardo [-@vaisey2016] find that cohort -- the year of a person's birth - explains much more variance than period -- the year of observation -- for most items in the General Social Survey, suggesting a strong impressionable years pattern.

@@

## Biological Mechanisms

The first potential mechanism for changes in attitude stability over time is rooted in the biology of the human brain and science of how people process information. This approach tends to suggest that human cognition works similar to Bayesian updating, in which humans weight all information they are exposed to equally, whether they encounter it during adolescence or middle age, and are exposed to information at a (more or less) constant rate [@bartels2014; @gopnik2020; @sherratt2018]. However, because people have already encountered a lot of information by the time they reach adulthood, each new piece of information encountered later has less of an impact on their overall dispositions [@bartels2014]. From this perspective, the distinctive thing about our early years is simply that they happen early, and they prove so influential because the incremental effect of new information diminishes as we age.

It is also possible that there is something unique about human cognition during these periods making people ascribe more weight to the information acquired within them than to information acquired at other points. Maturational arguments about the increased plasticity of human brains during our early years are consistent with this position, as our brains appear to exhibit more malleability during childhood and adolescence than later in life [@mata2015]. Supporting this model is evidence that early formative periods have been observed not just in people, but across species. In a study of social learning among birds, for instance, Aplin and colleagues [-@aplin2017] show that younger birds -- unlike their older counterparts -- were consistently able to retool their habits in light of new information.

@@

## Life Course and Social Roles

@@

Given the distinct social conditions of the early-adult phase of life, there are significant reasons to believe that the "impressionable years" are principally an effect of how societies choose to structure the life course or the interaction between these life course stages and biological mechanisms. Because of this, contemporary historical trends offer a window to disentangle the role of socio-cultural and biological or cognitive factors in explaining why we observe "impressionable years."

@@

First, the period between adolescence and "settling down" has been getting longer over time, and this might affect the patterns of attitudinal updating we expect to see for the same age groups in different cohorts. Buchmann [-@buchmann1989] finds that the diversification of life-trajectories has been accompanied by the postponement of the rituals associated with adulthood. In the United States, both the median age at first marriage and median age of first child birth has consistently risen since the mid-20th Century [@rosenfeld2009]. We also see higher rates of non-marital childbearing across all demographic groups [@cherlin2021]. Thus, the markers of the transition into adulthood seem to be occurring at later ages and potentially in different orders for more recent cohorts.

If the distinctive nature of "impressionable years" is principally driven by their unique social circumstances, then these shifts have the potential to extend these social circumstances into later ages. This, in turn, could have an effect on how long exploration, and therefore openness to durable change, lasts throughout the life-course. If this is the case, we should expect to see lower levels of attitude stability (and therefore more change) among age groups that, for earlier cohorts, displayed mainly stability. In other words, people who were born in the mid-20th century would have displayed mainly attitudinal stability by the time they reached their mid-twenties while more contemporary counterparts might still show cultural change when they reach that age. Specifically, we expect the attitude behavior of people aged 25-33 to more closely resemble middle-aged respondents in earlier cohorts [@krosnic1991; @alwin1991] but younger respondents in more recent cohorts. This produces our first hypothesis:

@@

## Intensifying Impressionable Years

A second trend in early adulthood relates to just how much exploration young adults can engage in during these "impressionable years." In his work, Rosenfeld [-@rosenfeld2007] argues that one of the processes underlying the rapid rise in same-sex and interracial marriages in the United States is that young adults are increasingly escaping the controlling gaze of their parents. When a young person lives with, or is financially dependent on, his or her parents, the latter have more say about their child's choice of partner. Social and cultural changes that have increased geographic mobility and independence from parental authority mean people in this life stage have been able to engage in courtship outside the gaze of their parents, making choices that would not have been sanctioned by them.

Rosenfeld's work suggests that the social conditions that shape where we spend our impressionable years might also shape how much exploration we can engage in. An increase in independence might entail more exploration, and greater ability to update attitudes as a result of this exploration. This assertion in turn engenders a prediction: we should see more attitudinal updating among young adults in recent years than among young adults in earlier eras, even if they have met similar life-course milestones.

@@

Reliability reflects not just the degree to which people make errors in their reporting of attitudes over time, but the baseline level of variance around a fixed attitude position that people express. In other words, we might ... if younger people bounce around
